### 2020-07-11 paper/sex_selection_spacing-ver4.tex
Clarify spell definition and model set-up
- Spell period from one parity to next birth or censoring
- Only have discussion of explanatory variables in one place
- Original proportionality discussion unclear
- Edit model set-up for clarity and introduce non-proportional part
more generically rather than with variables
- Footnote on endogenous sex composition should be when that variable
is introduced
- Specify that predicted birth intervals are in months

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are two problems with the standard approach in this setting.

First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up that captures both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, even without sex selection it is unlikely that characteristics, such as the sex
composition of previous births, have the same effects throughout the entire birth interval.
That is, however, the assumption behind the proportional hazard model, and if 
proportionality does not hold, the results are biased.
The proportionality assumption is especially problematic for higher-order birth 
intervals where there are substantial differences across groups both in the likelihood of 
progressing to the next birth and how soon couples want their next child if they are going 
to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, different groups have different levels of sex-selective abortion use and, thereby, 
birth spacing. 
Second, within a birth interval, a household's use of sex selection may vary and that
means that the effects of covariates vary as well.

I, therefore, use a non-proportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell---the period from nine months after one birth to the 
following birth or censoring.
The spells begin nine months after the previous birth since this is the earliest we should 
expect to observe a new birth.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or there are too few births for the method to work.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ indicates which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise constant baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous.
It is beyond the scope of this paper to develop a method for dealing with this issue.
}

@@

I divide the explanatory variables into two groups.
The first group consists of characteristics that the prior literature finds affect 
the spacing choice and the use of sex selection:
mother's education, sex composition of previous children, and area of residence.
The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether the household belongs to a scheduled tribe
or caste.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
subsample and then predict average birth spacing, sex ratio, and the probability of having 
a birth in that spell.

@@

The predicted parity progression probability is the likelihood of giving birth by the
end of the spell.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is to bias downward the differences in parity progression probabilities.
}
The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over, calculated as shown above.
For comparison, the graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

>>>>

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit---the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are two problems with the standard approach in this setting.

First and foremost, the use of sex selection means that the sex of the next child is 
no longer random and that the spacing to the birth of a boy will differ from the 
spacing to a girl. 
Therefore, I use a competing risk set-up that captures both the non-randomness of 
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, even without sex selection, it is unlikely that characteristics, such as 
the sex composition of previous births, have the same effects throughout the entire 
birth interval. 
The proportional hazard model requires that the hazard for any individual is a 
fixed proportion of the hazard for any other individual. 
That the effects are not constant violates that assumption, and the results from a 
proportional hazard model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals since there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, different groups have different levels of sex-selective abortion use and, thereby, 
birth spacing. 
Second, within a birth interval, a household's use of sex selection may vary, and that
means that the effects of covariates vary as well.

Therefore, I use a non-proportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete-time, non-proportional, competing risk hazard model with two 
exit states: either a boy or a girl is born. 
The unit of analysis is a spell---the period from one parity birth to the following 
birth or censoring. 
For estimation purposes, the spells begin nine months after the previous birth since 
this is the earliest we should expect to observe a new birth. 
Censoring can happen for three reasons: the survey takes place, sterilization of the 
woman or her husband, or imposed because of too few births for the method to work.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ indicates which event took place.

The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2.
 \label{eq:hazard}
\end{equation}
$D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.
$\mathbf{Z}$ is the non-proportional part, which includes the interactions between 
$D_j(t)$ and a set of explanatory variables and the interactions of those.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.

@@

I divide the explanatory variables into two groups.
The first group consists of characteristics that the prior literature finds affect 
the spacing choice and the use of sex selection:
mother's education, sex composition of previous children, and area of residence.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous.
It is beyond the scope of this paper to develop a method for dealing with this issue.
}
The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether the household belongs to a scheduled tribe
or caste.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
subsample and then predict average birth interval in months, sex ratio, and the 
probability of having a birth in that spell.

@@

The predicted parity progression probability is the likelihood of giving birth by the
end of the spell.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is to bias downward the differences in parity progression probabilities.
}
The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over, calculated as shown above.
For comparison, I also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
